Temporal ordering and registration of images in studies of developmental dynamics

### 2014-05-30 image_analysis_paper/pnas_draft2.tex
added yannis edits to paper

@@

\significancetext{Cross-sectional imaging studies are commonly used to study developmental dynamics in multiple organisms. In these studies, a representative developmental trajectory is reconstructed from snapshots of different embryos, each of which has been arrested at a different point of its development. One of the first steps in this reconstruction involves registration and temporal ordering of images. Here we show that both of these tasks can be automated and combined in a single step using recently developed advances in data mining. To illustrate this approach, we temporally order and register several datasets collected in cross-sectional studies of cell signaling in the early fruit fly embryo.}

@@

Imaging studies provide unique insights into the dynamics of developmental processes. 
%
A large number of imaging studies are based on a cross-sectional experimental design, where developmental progress is arrested in a population of embryos, each of which is at a different point along its developmental trajectory. 
%
The goal is then to temporally order the data to reconstruct representative developmental dynamics from snapshots provided by the collection of embryos.
%
Images of different biological samples must first be registered before they can be temporally ordered.
%
When such data sets are large, noisy, and/or if the developmental changes are subtle, these tasks can be difficult to perform manually.
%
We present an automated approach to registering and temporally ordering data from such cross-sectional studies.
%
The approach is based on vector diffusion maps, a broadly applicable manifold learning technique that does not require {\it a priori} knowledge of image features or developmental dynamics.
%
%Furthermore, these techniques can separate multiple trajectories within the same data set.
%
We demonstrate the utility of these methods using a collection of images from a cross-sectional study of cell signaling during dorsoventral patterning of the {\it Drosophila} embryo.
\end{abstract}

@@

\dropcap{E}xperimental studies of developmental dynamics fall in two broadly defined categories: longitudinal and cross-sectional \cite{diggle2002analysis}.
%
In longitudinal studies, developmental progress is monitored over time in the same embryo \cite{roelens2013live, keller2013imaging}.
%
In a cross-sectional study, developmental dynamics is reconstructed from multiple embryos, each of which contributes only a snapshot of a chemical or morphological process along its developmental trajectory \cite{jaeger2004dynamic, fowlkes2008quantitative}.
%
Here we focus on cross-sectional studies, which have a time-honored history and still present the only option for most organisms.
%
In a typical cross-sectional study, a group of embryos is fixed using a procedure that arrests their development and stained with chemicals that visualize a handful of cellular processes.
%
Fixed embryos are then imaged using any given number of microscopy techniques.
%
%Recent advances in physical manipulation and imaging of embryos produce rapidly increasing volumes of cross-sectional data, in which every embryo is observed at a different developmental time point and at a different geometric orientation.
%
Importantly, the ``age'' of any given embryo arrested in its development is not quantitatively known; typically what is known is
a certain time window to which a collection of embryos belongs \cite{ng2012large, richardson2014emage, castro2009automatic}.
%to high accuracy.
%
In order to recover the developmental dynamics from such data sets, snapshots of different embryos must be spatially aligned or {\em registered} to factor out the relevant geometric symmetries (e.g., translations and rotations), and then ordered in time.
%
We show how existing dimensionality reduction algorithms can automate {\it and combine} both of these tasks.

@@

\begin{figure}[t]
\includegraphics[width=8.7cm]{fig1}
\caption{Caricature illustrating the tasks of image registration and temporal ordering. {\xfigtextfontit (A)} Images of ``samples'', each in a different orientation and a different stage of development. {\xfigtextfontit (B)} Registered and ordered samples. For this caricature, the registration and ordering is straightforward because the data set is small and the developmental changes are easy to recognize.}
%\label{fig:fish}
\customlabel{fig:fish}{1}
\customlabel{subfig:fish_unordered}{\ref{fig:fish}{\it A}}
\customlabel{subfig:fish_ordered}{\ref{fig:fish}{\it B}}
\end{figure}

@@

Image registration has been studied in a variety of contexts \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, 
and most registration algorithms are based on landmarks, such as the head and fins in \fig~\ref{fig:fish}.
%
In contrast, our approach requires no such information. 
%
It is based on a manifold learning algorithm \cite{singer2012vector} which simultaneously addresses the problems of registration and temporal ordering. 
%
This algorithm is one of many nonlinear dimensionality reduction techniques that have been developed over the past decade \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}. 
%
This technique has been used for data mining in
applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules to face recognition, and is sufficiently general to be applicable to a wide variety of biological imaging studies. 
%
Here the algorithm is adapted for simultaneous registration and temporal ordering of images from cross-sectional studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories.
%
To illustrate our approach, we analyze three data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of developmental dynamics \cite{jaeger2012drosophila}.
%
%The first data set, shown in Figure~\ref{subfig:raw_data1}, is relatively simple and will allow us to both illustrate and validate this approach.
%
%In the second data set, shown in Figure~\ref{subfig:raw_data2}, the dynamics are significantly more complex.
%
%Here, we will show that we can recover a trajectory that is qualitatively consistent with current knowledge.
%
%The third set consists of data collected from both wild type and mutant embryos, where joint processing of the data will help to visualize the different sources of variability along distinct developmental paths.

@@

\section{Results and Discussion}

@@

%%%somewhere I want a sentence that says this is for general symmetry groups of some kind, and HERE it is for..... (maybe in SI)
Angular synchronization uses pairwise alignment information to register a set of images in a globally consistent way.
%
A schematic illustration of angular synchronization is shown in \fig~\ref{subfig:synch1}, where each image is represented as a vector, and the goal is to align the set of vectors.
%
We first compute the angles needed to align pairs of vectors (or images).  
%
In general, this requires no template function \cite{ahuja2007template} or image landmarks \cite{ian1998statistical};
however, when the images are noisy, some pairwise alignments may be inaccurate.
%
Using the alignment angles between all pairs of vectors, angular synchronization finds the set of rotation angles (one angle for each vector) that is most consistent with all pairwise measurements (see {\it SI Materials and Methods}).
%
In this schematic, registration via angular synchronization is trivial as the pairwise measurements contain no noise.
%
However, the algorithm can successfully register data sets even when many of the pairwise measurements are corrupted by noise \cite{singer2011angular}.
%

%
After removing variability due to translations and rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional manifold that parameterizes most of the remaining variability in the data.
%%%
%%%YGK feels like repeats
%%%
%
Such a manifold can be discovered using diffusion maps \cite{coifman2005geometric}, a nonlinear dimensionality reduction technique that uncovers an intrinsic parameterization of data that lies on a low-dimensional, perhaps nonlinear, manifold in high-dimensional space.
%
The idea is illustrated in \fig~\ref{subfig:dmaps1}, where the data are two-dimensional points which lie on a one-dimensional nonlinear curve.
%
We use {\it local} information about the data to find a parameterization which respects the underlying manifold geometry: we want points which are close in high-dimensional space (e.g., images which look similar) to be close in our parameterization.
%
This idea of locality is denoted by the edges in \fig~\ref{subfig:dmaps1}.
%
Data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
%
The color in \fig~\ref{subfig:dmaps2} depicts the one-dimensional parameterization or ordering of the data that we can detect visually.
%
In our working examples, each data point will be much higher dimensional (e.g, an image with many pixels), and so we cannot extract this low-dimensional structure visually.
%
Instead, we will use diffusion maps to automatically uncover a parameterization of our high-dimensional data (see \SI).
%
Diffusion maps also generalizes from one-dimensional nonlinear curves to higher-dimensional manifolds.
%
Thus, when our experiments have additional sources of variability (such as distinct populations of embryos) in addition to developmental dynamics, diffusion maps can provide an informative organization of our data which highlights both of these variability sources.
%We will assume that our data are approximately one-dimensional, and that this dimension is parameterized by time, such that ordering our data along this main dimension/direction will temporally order our data.

Both angular synchronization and diffusion maps use pairwise information to uncover global structure in the data.
%
Angular synchronization uses pairwise alignment information to find translations and rotations which are globally consistent, and diffusion maps uses pairwise distances to find parameterizations which are globally informative.
%
Vector diffusion maps combines both steps into a single computation (see \SI), simultaneously addressing the tasks of registration and temporal ordering.
%
%The algorithm uses information about pairwise alignments {\em and} pairwise distances to find a global parameterization of the data, while registering images which are ``close'' .
%
%It is thus perfectly suited for our task of registering and temporally ordering cross-sectional imaging data.
%

\begin{figure}[t]
\includegraphics[width=8.7cm]{fig2}
\caption{Schematic illustrating angular synchronization and diffusion maps. {\xfigtextfontit (A)} Set of vectors, each in a different orientation. The pairwise alignment angles are indicated. {\xfigtextfontit (B)} The vectors from {\xfigtextfontit A}, each rotated about their midpoint so that the set is globally aligned. Note that the chosen rotation angles are consistent with the pairwise alignments in {\xfigtextfontit A}: the differences between a pair of angles in {\xfigtextfontit B} is the same as the pairwise angle in {\xfigtextfontit A}. {\xfigtextfontit (C)} Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see {\xfigtextfontit \SI}), so that close data points are connected by dark (``stronger'') edges. {\xfigtextfontit (D)} The data in {\xfigtextfontit C}, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure. The color intensity is monotonic with the arclength, thus parameterizing the curve.}
%\label{fig:schematics}
\customlabel{fig:schematics}{2}
\customlabel{subfig:synch1}{\ref{fig:schematics}{\it A}}
\customlabel{subfig:synch2}{\ref{fig:schematics}{\it B}}
\customlabel{subfig:dmaps1}{\ref{fig:schematics}{\it C}}
\customlabel{subfig:dmaps2}{\ref{fig:schematics}{\it D}}
\end{figure}

\begin{figure*}[t]
\includegraphics{fig3}
\caption{{\xfigtextfontit Drosophila} during cellularization. {\xfigtextfontit (A)} Images of {\xfigtextfontit Drosophila} embryos collected during cellularization. Each image is of a different embryo in a different rotational and translational orientation. {\xfigtextfontit (B)} Images from {\xfigtextfontit A}, now registered and ordered using vector diffusion maps. The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom.}
\customlabel{fig:data1}{3}
\customlabel{subfig:raw_data1}{\ref{fig:data1}{\it A}}
\customlabel{subfig:ordered_data1}{\ref{fig:data1}{\it B}}
\customlabel{subfig:averaged_data1}{\ref{fig:data1}{\it C}}
\end{figure*}

@@

As a first illustration of using vector diffusion maps to temporally order and register images, we analyze a data set where the correct registration and temporal order is known {\it a priori}.
%
This data set is obtained by fluorescent imaging of {\it Drosophila} embryos during the third hour after fertilization.
%
By this stage of development, sequential divisions of the fertilized egg have generated a system where most of the nuclei are arranged in a monolayer under the common plasma membrane \cite{foe1983studies}.
%
\fig~\ref{subfig:raw_data1} shows optical cross-sections of each embryo, with nuclei (gray in the figure) labeled by DAPI, a DNA stain.
%
During the third hour of development, the spatially uniform two-dimensional arrangement of nuclei is patterned by the concentration profiles of chemical signals that control cell differentiation.
%
One of these signals is provided by the nuclear localization of the transcription factor Dorsal (Dl, green in the figure), which subdivides the early embryo into the territories that give rise to the muscle, nerve, and skin tissues.
%
Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, red in the figure), an enzyme that specifies a subset of neuronal cells.
%

@@

For this data set, the quality of the temporal ordering provided by vector diffusion maps can be assessed by comparing with the ordering obtained through an independent approach.
%
During the third hour of development, the embryo cellularizes: in this process, nuclei located under a common plasma membrane are enclosed by lateral membranes which grow inward, separating the nuclei into individual cells.
%
The highly reproducible kinetics of lateral membrane growth can be used to estimate the age of each embryo to within 2--3 minutes \cite{figard2013plasma} (see \fig~S2).
%
The Spearman rank correlation coefficient between the orderings obtained based on the progress of cellularization and vector diffusion maps is 0.91, indicating that vector diffusion maps can accurately recover the temporal order.
%
Thus, we have validated a dimensionality reduction approach to temporal ordering and registration on a data set where image orientation and ordering can be assessed independently.

@@

%As the images in Figure~\ref{fig:data1} are relatively simple, other techniques could have been used to temporally order them.
%
Because the dynamics of the data set shown in \fig~\ref{fig:data1} are relatively simple, effectively consisting of a single mode (the two dpERK peaks) growing in time, established techniques such as principal component analysis (PCA) \cite{shlens2005tutorial} would uncover most of the meaningful structure in the data, and projection onto the first principal component would be sufficient to order the data in time (see \fig~S7).
%
However, developmental dynamics are often significantly more complex, and nonlinear dimensionality reduction techniques are required to extract meaningful structure in the data.
%

\fig~\ref{subfig:raw_data2} shows a richer data set of $108$ images of {\it Drosophila} embryos.
%
These images cover a thirty minute time interval and span late cellularization through gastrulation, when the two-dimensional sheet of newly formed cells begins to deform into three-dimensional structures which will form the future organs \cite{leptin2005gastrulation}.
%
As before, each image is an optical cross-section of a vertically oriented embryo fixed at a different developmental time.
%
The nuclei (gray) are again labeled with a DNA marker.
%
However, during this developmental time period, the nuclei are no longer uniformly arranged around the periphery of the embryo.
%
Instead of Dl, which is difficult to visualize at this point, embryos were stained with the antibody that recognizes Twist (Twi, shown in green), a transcription factor that is activated by high levels of Dl, specifying the cells of the future muscle tissue.
%
dpERK is again shown in red.

%
During this window, the ventral furrow is formed, where the ventral side of the embryo buckles towards the center of the embryo, bringing in the future muscle cells.
%
dpERK appears as two lateral peaks at the ventral side of the embryo.
%
These two peaks merge together during invagination, eventually forming (together with Twi) a characteristic ``omega'' shape.
%
During this time window, germband extension causes cells from the ventral side to move towards the posterior pole of the embryo, and then wrap around to the dorsal side \cite{leptin2005gastrulation}.
%
At the end of this process, cells which were originally on the ventral and posterior side of the embryo find themselves on the dorsal side, causing similar ``omega'' shaped patterns to arise on the dorsal side; these patterns are most readily seen in the last image of \fig~\ref{subfig:averaged_data2}.

\begin{figure*}[t]
\includegraphics{fig4}
\caption{{\xfigtextfontit Drosophila} during gastrulation. {\xfigtextfontit (A)} Images of gastrulating {\xfigtextfontit Drosophila} embryos. Each image is of a different embryo in a different rotational and translational orientation. {\xfigtextfontit (B)} Data from {\xfigtextfontit A}, registered and ordered using vector diffusion maps. {\xfigtextfontit (C)} Average trajectory of the images shown in {\xfigtextfontit B}. Each image is a (Gaussian-weighted) average of the images in {\xfigtextfontit B}. }
\customlabel{fig:data2}{4}
\customlabel{subfig:raw_data2}{\ref{fig:data2}{\it A}}
\customlabel{subfig:ordered_data2}{\ref{fig:data2}{\it B}}
\customlabel{subfig:averaged_data2}{\ref{fig:data2}{\it C}}
\end{figure*}

Visually, the dynamics in this data set are much more complicated, and only 22\% of the variability in the images is captured by the first principal component.
%
Simple time-ordering based on this linear projection is therefore not possible (see \fig~S8).
%
%Even when projecting on several leading principal components (see SI), a nonlinear approach, such as vector diffusion maps, is required to temporally order the data.
%
\fig~\ref{subfig:ordered_data2} shows the images in \fig~\ref{subfig:raw_data2}, now registered and ordered using vector diffusion maps \cite{singer2012vector}.
%
Visual inspection confirms this ordering is consistent with the known developmental dynamics outlined above.
%
However, assimilating the dynamics even from this concatenation of images is difficult for two reasons, even after registration and ordering.
%
First, the sheer number of images makes visual processing of the entire data set nontrivial, and
second, the ordered data is not entirely smooth due to inter-embryo variability.
%
To highlight the developmental dynamics revealed by vector diffusion maps, we applied a Gaussian filter to the trajectory of registered and ordered images to obtain an averaged developmental path.  
%
Sequential snapshots of this averaged trajectory, shown in \fig~\ref{subfig:averaged_data2}, serve as a summary of a stereotypic developmental dynamics reconstructed from cross-sectional data.
%
Thus, vector diffusion maps can accomplish the image registration and ordering tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks (e.g., fins) and without {\it a priori} knowledge of developmental dynamics (e.g., correlation of age with body size).

@@

\section{Microscopy}
%
Nikon A1-RS scanning confocal microscope, and the Nikon 60X Plan-Apo oil objective was used to image embryos. Embryos were collected, stained, and imaged together under the same microscope setting. End-on imaging was performed by using the microfluidics device described previously. Images were collected at the focal plane $\sim$90~$\mathrm{\mu m}$ from the posterior pole of an embryo.

>>>>

\customlabel{fig:DV_view}{S1}
\customlabel{fig:membrane_curve}{S2}
\customlabel{fig:raw_data3}{S3}
\customlabel{fig:1d_example}{S4}
\customlabel{fig:spheres}{S5}
\customlabel{fig:eigenvalues}{S6}
\customlabel{fig:PCA_data1}{S7}
\customlabel{fig:PCA_data2}{S8}

@@

\significancetext{Cross-sectional imaging studies are commonly used to study developmental dynamics in multiple organisms. 
%
In these studies, a representative developmental trajectory is reconstructed from snapshots of different embryos, each of which has been arrested at a different point of its development. 
%
One of the first steps in this reconstruction involves registration and temporal ordering of images. 
%
Here we show that both of these tasks can be automated and combined in a single step using recent advances in data mining. 
%
To illustrate this approach, we temporally order and register several datasets collected in cross-sectional studies of cell signaling in the early fruit fly embryo.}

@@

Imaging studies provide unique insights into the dynamics of developmental processes. 
%
A large number of imaging studies are based on a cross-sectional experimental design, where developmental progress is arrested in a population of embryos, each of which is at a different point along its developmental trajectory. 
%
The goal is then to temporally order the data to reconstruct representative developmental dynamics from snapshots provided by the collection of embryos.
%
Images of different biological samples must first be registered before they can be temporally ordered.
%
When such data sets are large, noisy, and/or if the developmental changes are subtle, these tasks can be difficult to perform manually.
%
We present an automated approach to simultaneously register and temporally order data from such cross-sectional studies.
%
The approach is based on vector diffusion maps, a broadly applicable manifold learning technique that does not require {\it a priori} knowledge of image features or a parametric model of the developmental dynamics.
%
%Furthermore, these techniques can separate multiple trajectories within the same data set.
%
We use this method to extract developmental dynamics from collections of images from a cross-sectional study of cell signaling during dorsoventral patterning of the {\it Drosophila} embryo.
\end{abstract}

@@

\dropcap{E}xperimental studies of developmental dynamics fall in two broadly defined categories: longitudinal and cross-sectional \cite{diggle2002analysis}.
%
In longitudinal studies, developmental progress is monitored over time in the same embryo \cite{roelens2013live, keller2013imaging}.
%
In a cross-sectional study, developmental dynamics is reconstructed from multiple embryos, each of which contributes only a snapshot of a chemical or morphological process along its developmental trajectory \cite{jaeger2004dynamic, fowlkes2008quantitative}.
%
Here we focus on cross-sectional studies, which have a time-honored history and still present the only option for most organisms.
%
In a typical cross-sectional study, a group of embryos is fixed using a procedure that arrests their development and stained with chemicals that visualize a handful of cellular processes.
%
Fixed embryos are then imaged using any given number of microscopy techniques.
%
%Recent advances in physical manipulation and imaging of embryos produce rapidly increasing volumes of cross-sectional data, in which every embryo is observed at a different developmental time point and at a different geometric orientation.
%
Importantly, the ``age'' of any given embryo arrested in its development is not quantitatively known; typically what is known is
a certain time window to which a collection of embryos belongs \cite{ng2012large, richardson2014emage, castro2009automatic}.
%to high accuracy.
%
In order to recover the developmental dynamics from such data sets, snapshots of different embryos must first be spatially aligned or {\em registered} to factor out the relevant geometric symmetries (e.g., translations and rotations), and then ordered in time.
%
We show how recently developed dimensionality reduction algorithms can automate {\it and combine} both of these tasks.

@@

\begin{figure}[t]
\includegraphics[width=8.7cm]{fig1}
\caption{Caricature illustrating the tasks of image registration and temporal ordering. {\xfigtextfontit (A)} Images of ``samples'', each in a different orientation and a different stage of development. {\xfigtextfontit (B)} Registered and ordered samples. For this caricature, the registration and ordering is straightforward because the data set is small and the developmental changes are easy to recognize visually.}
%\label{fig:fish}
\customlabel{fig:fish}{1}
\customlabel{subfig:fish_unordered}{\ref{fig:fish}{\it A}}
\customlabel{subfig:fish_ordered}{\ref{fig:fish}{\it B}}
\end{figure}

@@

Image registration has been studied in a variety of contexts \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, 
and most registration algorithms are based on landmarks, such as the head and fins in \fig~\ref{fig:fish}.
%
In contrast, our approach requires no such information. 
%
It is based on a manifold learning algorithm (vector diffusion maps, VDM \cite{singer2012vector}) which simultaneously addresses the problems of registration and temporal ordering. 
%
This algorithm is one of several influential nonlinear dimensionality reduction techniques that have been developed over the past decade \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}. 
%
The technique has been used for data mining in
applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules to face recognition, and is sufficiently general to be applicable to a wide variety of biological imaging studies. 
%
Here the algorithm is adapted for simultaneous registration and temporal ordering of images from cross-sectional studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories.
%
To illustrate our approach, we analyze three data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of developmental dynamics \cite{jaeger2012drosophila}.
%
%The first data set, shown in Figure~\ref{subfig:raw_data1}, is relatively simple and will allow us to both illustrate and validate this approach.
%
%In the second data set, shown in Figure~\ref{subfig:raw_data2}, the dynamics are significantly more complex.
%
%Here, we will show that we can recover a trajectory that is qualitatively consistent with current knowledge.
%
%The third set consists of data collected from both wild type and mutant embryos, where joint processing of the data will help to visualize the different sources of variability along distinct developmental paths.

@@

\section{Results}

@@

%%%somewhere I want a sentence that says this is for general symmetry groups of some kind, and HERE it is for..... (maybe in SI)
Angular synchronization uses pairwise alignment information to register a set of images in a globally consistent way.
%
A schematic illustration of angular synchronization is shown in \fig~\ref{subfig:synch1}, where each image is represented as a vector, and the goal is to align the set of vectors.
%
We first compute the angles needed to align pairs of vectors (or images).  
%
In general, this requires no template function \cite{ahuja2007template} or image landmarks \cite{ian1998statistical}.
%
Using the alignment angles between all pairs of vectors, angular synchronization finds the set of rotation angles (one angle for each vector) that is most consistent with {\it all} pairwise measurements (see {\it SI Materials and Methods}); this is illustrated in \fig~\ref{subfig:synch2}.
%
In this schematic, registration via angular synchronization is trivial, as the pairwise measurements contain no noise.
%
However, the algorithm can successfully register data sets even when many of the pairwise measurements are inaccurate \cite{singer2011angular}.
%

%
After removing variability due to translations and rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional manifold that parameterizes most of the remaining variability in the data.
%%%
%%%YGK feels like repeats
%%%
%
Such a manifold can be discovered using diffusion maps \cite{coifman2005geometric}, a nonlinear dimensionality reduction technique that uncovers an intrinsic parametrization of data that lies on a low-dimensional, perhaps nonlinear, manifold in high-dimensional space.
%
The idea is illustrated in \fig~\ref{subfig:dmaps1}, where the data are two-dimensional points which lie on a one-dimensional nonlinear curve.
%
We use {\it local} information about the data to find a parametrization  which respects the underlying manifold geometry: we want points which are close in high-dimensional space (e.g., images which look similar) to be close in our parametrization.
%
This idea of locality is denoted by the edges in \fig~\ref{subfig:dmaps1}.
%
Data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
%
The color in \fig~\ref{subfig:dmaps2} depicts the one-dimensional parametrization or ordering of the data that we can detect visually.
%
In our working examples, each data point will be much higher dimensional (e.g., a pixelated image), and so we cannot extract this low-dimensional structure visually.
%
Instead, we will use diffusion maps to automatically uncover a parametrization of our high-dimensional data (see \SI).
%
Diffusion maps generalizes directly from one-dimensional nonlinear curves to higher-dimensional manifolds.
%
Thus, when our experiments have additional sources of variability (such as distinct populations of embryos) in addition to developmental dynamics, diffusion maps can provide an informative organization of our data which highlights both of these variability sources.
%We will assume that our data are approximately one-dimensional, and that this dimension is parameterized by time, such that ordering our data along this main dimension/direction will temporally order our data.

Both angular synchronization and diffusion maps use pairwise information to uncover global structure in the data.
%
Angular synchronization uses pairwise alignment information to find translations and rotations which are globally consistent, and diffusion maps uses pairwise distances to find global parametrizations which are (geometrically) informative.
%
Vector diffusion maps combines both steps into a single computation (see \SI), simultaneously addressing the tasks of registration and temporal ordering.
%
%The algorithm uses information about pairwise alignments {\em and} pairwise distances to find a global parametrization of the data, while registering images which are ``close'' .
%
%It is thus perfectly suited for our task of registering and temporally ordering cross-sectional imaging data.
%

\begin{figure}[t]
\includegraphics[width=8.7cm]{fig2}
\caption{Schematic illustrating angular synchronization and diffusion maps. {\xfigtextfontit (A)} Set of vectors, each in a different orientation. The pairwise alignment angles are indicated. {\xfigtextfontit (B)} The vectors from {\xfigtextfontit A}, each rotated about their midpoint so that the set is globally aligned. Note that the chosen rotation angles are consistent with the pairwise alignments in {\xfigtextfontit A}: the differences between a pair of angles in {\xfigtextfontit B} is the same as the pairwise angle in {\xfigtextfontit A}. {\xfigtextfontit (C)} Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see {\xfigtextfontit \SI}), so that close data points are connected by darker (``stronger'') edges. {\xfigtextfontit (D)} The data in {\xfigtextfontit C}, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure. The color intensity is monotonic with the perceived curve arclength, thus parameterizing the curve.}
%\label{fig:schematics}
\customlabel{fig:schematics}{2}
\customlabel{subfig:synch1}{\ref{fig:schematics}{\it A}}
\customlabel{subfig:synch2}{\ref{fig:schematics}{\it B}}
\customlabel{subfig:dmaps1}{\ref{fig:schematics}{\it C}}
\customlabel{subfig:dmaps2}{\ref{fig:schematics}{\it D}}
\end{figure}

\begin{figure*}[t]
\includegraphics{fig3}
\caption{{\xfigtextfontit Drosophila} embryos during cellularization. {\xfigtextfontit (A)} Images of embryos collected during cellularization. Each image is of a different embryo arrested at a different developmental time, and in a different rotational and translational orientation. {\xfigtextfontit (B)} Images from {\xfigtextfontit A}, now registered and ordered using vector diffusion maps. The images are ordered from left to right and from top to bottom, so that the first image appears in the top left position of the array, and the last image appears in the bottom right position. The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom.}
\customlabel{fig:data1}{3}
\customlabel{subfig:raw_data1}{\ref{fig:data1}{\it A}}
\customlabel{subfig:ordered_data1}{\ref{fig:data1}{\it B}}
\customlabel{subfig:averaged_data1}{\ref{fig:data1}{\it C}}
\end{figure*}

@@

As a first illustration of using vector diffusion maps to temporally order and register images, we analyze a data set where the correct registration and temporal order is known {\it a priori}.
%
This data set is obtained by fluorescent imaging of {\it Drosophila} embryos during the third hour after fertilization.
%
By this stage of development, sequential divisions of the fertilized egg have generated a system where most of the nuclei are arranged in a monolayer under the common plasma membrane \cite{foe1983studies}.
%
\fig~\ref{subfig:raw_data1} shows optical cross-sections of each embryo, with nuclei (gray in the figure) labeled by DAPI, a DNA stain.
%
During the third hour of development, this arrangement of nuclei is patterned by the concentration profiles of chemical signals that control cell differentiation.
%
One of these signals is provided by the nuclear localization of the transcription factor Dorsal (Dl, green in the figure), which subdivides the early embryo into the territories that give rise to the muscle, nerve, and skin tissues.
%
Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, red in the figure), an enzyme that specifies a subset of neuronal cells.
%

@@

For this data set, the quality of the temporal ordering provided by vector diffusion maps can be assessed by comparing with the ordering obtained through an independent approach.
%
During the third hour of development, the embryo cellularizes: in this process, nuclei located under a common plasma membrane are enclosed by lateral membranes which grow inward, separating the nuclei into individual cells.
%
The highly reproducible kinetics of lateral membrane growth can be used to estimate the age of each embryo to within 2--3 minutes \cite{figard2013plasma} (see \fig~\ref{fig:membrane_curve}).
%
The Spearman rank correlation coefficient between the orderings obtained based on the progress of cellularization and vector diffusion maps is 0.91, indicating that vector diffusion maps can accurately recover the temporal order.
%
Thus, we have validated a dimensionality reduction approach to temporal ordering and registration on a data set where image orientation and ordering can be assessed independently.

@@

%As the images in Figure~\ref{fig:data1} are relatively simple, other techniques could have been used to temporally order them.
%
Because the dynamics of the data set shown in \fig~\ref{fig:data1} are relatively simple, effectively consisting of a single mode (the two dpERK peaks) growing in time, established techniques such as principal component analysis (PCA) \cite{shlens2005tutorial} would uncover most of the meaningful structure in the data, and projection onto the first principal component would be sufficient to order the data in time (see \fig~\ref{fig:PCA_data1}).
%
However, developmental dynamics are often significantly more complex, and nonlinear dimensionality reduction techniques are required to extract meaningful structure.
%
\fig~\ref{subfig:raw_data2} shows a richer data set of $108$ images of {\it Drosophila} embryos.
%
These images cover a thirty minute time interval and span late cellularization through gastrulation, when the two-dimensional sheet of newly formed cells begins to deform into three-dimensional structures which will form the future organs \cite{leptin2005gastrulation}.
%
As before, each image is an optical cross-section of a vertically oriented embryo fixed at a different developmental time.
%
The nuclei (gray) are again labeled with a DNA marker.
%
However, during this developmental time period, the nuclei are no longer uniformly arranged around the periphery of the embryo.
%
Instead of Dl, which is difficult to visualize at this point, embryos were stained with the antibody that recognizes Twist (Twi, shown in green), a transcription factor that is activated by high levels of Dl, specifying the cells of the future muscle tissue.
%
dpERK is again shown in red.

%
During this time window the ventral furrow is formed, where the ventral side buckles towards the center of the embryo, bringing in the future muscle cells.
%
dpERK appears as two lateral peaks at the ventral side of the embryo.
%
These two peaks merge together during invagination, eventually forming (together with Twi) a characteristic ``omega'' shape.
%
Germband extension then causes cells from the ventral side to move towards the posterior pole of the embryo, and then wrap around to the dorsal side \cite{leptin2005gastrulation}.
%
At the end of this process, cells which were originally on the ventral and posterior side of the embryo find themselves on the dorsal side, causing similar ``omega'' shaped patterns to arise on the dorsal side; these patterns are most readily seen in the last image of \fig~\ref{subfig:averaged_data2}.

\begin{figure*}[t]
\includegraphics{fig4}
\caption{{\xfigtextfontit Drosophila} embryos during gastrulation. {\xfigtextfontit (A)} Images of gastrulating embryos. Each image is of a different embryo arrested at a different developmental time, and in a different translation and rotational orientation. {\xfigtextfontit (B)} Data from {\xfigtextfontit A}, registered and ordered using vector diffusion maps. The images are ordered from left to right and from top to bottom, so that the first image appears in the top left position of the array, and the last image appears in the bottom right position. {\xfigtextfontit (C)} A representative ``developmental path'' from local averaging of the images. Each image is a (Gaussian-weighted) average of the images in {\xfigtextfontit B}. }
\customlabel{fig:data2}{4}
\customlabel{subfig:raw_data2}{\ref{fig:data2}{\it A}}
\customlabel{subfig:ordered_data2}{\ref{fig:data2}{\it B}}
\customlabel{subfig:averaged_data2}{\ref{fig:data2}{\it C}}
\end{figure*}

Visually, the dynamics in this data set are much more complicated, and only 22\% of the variability in the images is captured by the first principal component.
%
Simple time-ordering based on this linear projection is therefore not possible (see \fig~\ref{fig:PCA_data2}).
%
%Even when projecting on several leading principal components (see SI), a nonlinear approach, such as vector diffusion maps, is required to temporally order the data.
%
\fig~\ref{subfig:ordered_data2} shows the images registered and ordered using vector diffusion maps \cite{singer2012vector}, 
and visual inspection confirms this ordering is consistent with the known developmental dynamics outlined above.
%
However, assimilating the dynamics even from this concatenation of images is difficult for two reasons, even after registration and ordering.
%
First, the sheer number of images makes visual processing of the entire data set nontrivial;
and second, the ordered data is not entirely smooth due to inter-embryo variability.
%
To highlight the developmental dynamics revealed by vector diffusion maps, we applied a Gaussian filter to the trajectory of registered and ordered images to obtain an averaged developmental path (see also \cite{kemelmacher2011exploring}).  
%
Sequential snapshots of this averaged trajectory, shown in \fig~\ref{subfig:averaged_data2}, serve as a summary of the stereotypic developmental dynamics reconstructed from cross-sectional data.
%
Thus, vector diffusion maps can accomplish the image registration and ordering tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks (e.g., fins) and without {\it a priori} knowledge of developmental dynamics (e.g., correlation of age with body size).

@@

\section{Microscopy}
%
Nikon A1-RS scanning confocal microscope, and the Nikon 60X Plan-Apo oil objective was used to image embryos. Embryos were collected, stained, and imaged together under the same microscope setting. End-on imaging was performed by using the microfluidics device described previously. Images were collected at the focal plane $\sim$90~$\mathrm{\mu m}$ from the posterior pole of an embryo (see \fig~\ref{fig:DV_view}).
